---
title: "Introduction to Data Reading and Tidying"
author: Abhinav Anand, IIMB
date: '`r format(Sys.time(), "%Y/%m/%d")`' #current date

output:
  pdf_document:
    keep_tex: true

fontsize: 11pt
documentclass: article
geometry: margin = 1.5in

linkcolor: blue
urlcolor: red
citecolor: magenta

citation_package: natbib
bibliography: Working_Paper.bib

header-includes:
   - \linespread{1.25}


---

```{r setup, eval=T, message=FALSE, warning=T, include=FALSE, echo=F}

library(tidyverse)
library(rmarkdown)
library(knitr)
library(xml2)
library(gapminder)

knitr::opts_chunk$set(echo = T, 
                      warning = T, 
                      message = F, 
                      eval = T, 
                      include = T
                      )
```

# Setup

The packages `readr` and `tidyr` need to be installed prior to running the commands below. They are included in `tidyverse` by default. To install, type in the console `install.packages(c("readr", "tidyr"))`; or equivalently `install.packages("tidyverse")`


# Reading and Parsing Data Files

The following discussion assumes that all data files referenced are in the same folder as the R codes.

## Reading Plain-Text Files (`.csv`, `.tsv` etc.)

We will be working with the following set of files to illustrate the ideas regarding reading real-life, empirical data files.

```{r data_files}

file_fin_risk <- "FMC_T4_read_file_fin_risk.csv"
file_gdppc <- "FMC_T4_read_file_gdppc.csv"
file_US_corp_spread <- "FMC_T4_read_file_US_corp_spread.csv"



```

While we may read files in formats such as `.xls`, `.xlsx` etc. ("excel files") in R by using the package `readxl`, it is advised by many writers to convert such files into plain-text `.csv` format (comma separated format) and then open them by the `readr` package functions.

### `read_csv()`

`read_csv()` reads `.csv` files. For semicolon separated files, `read_csv2()` function is used. 

```{r data_read, message=T}

(fin_risk <- readr::read_csv(file_fin_risk)) #file path


```

`read_csv` uses the first row as the column names of data. If however, we know this to not be true (sometimes there are a few lines of metadata at the top of the file) we can instruct `read_csv` to refrain from such behavior.

```{r data_read_skip}

(US_corp_spread <- readr::read_csv(file_US_corp_spread))

(US_corp_spread_skip <- readr::read_csv(file_US_corp_spread, 
                                        skip = 10)
  )

```

### Notes

1. When the data file does not have column names we can use `col_names = FALSE` to tell `read_csv()` not to treat the first row as headings, and instead label them sequentially from $X1$ to $Xn$.

2. While base R has the classic `read.csv()` function to read `.csv` files, usage of `read_csv()` is encouraged since the latter is said to be around 10 times faster than the former. This is critical when file sizes become large. Additionally, the files are read as tibbles and hence retain their readability, flexibility and reproduceability. 

3. Excel files can be read with `readxl()`. Files in formats foreign to R, such as Stata files (`.dta`) can be read using the tidyverse package `haven`.

4. R can also write dataframes into a `.csv` file by use of the command `write_csv()`.


# Tidying Data

```{r data_read_alt, message=T}

(gdppc <- readr::read_csv(file_gdppc)) #which format?

head(fin_risk) #which format?

```

## The Tidy Format

The tidy format has three characteristics:

1. Each variable is a column
2. Each observation is a row
3. Each value is a cell

Not all formats of data are equally good for analysis. For the tidyverse, the best format to work with is the "tidy" format. `dplyr`, `ggplot2` and all the other packages in the tidyverse are designed to work best with tidy data.

`fin_risk` is a tidy dataset, `gdppc` is not.

```{r data_tidy_read}

(fin_risk_tidy <- fin_risk %>%
  dplyr::rename(risk_foreign = 
                  `Risk Points for Foreign Debt as a % of GDP`) %>%
  dplyr::rename(risk_exchange = 
                  `Risk Points for Exchange Rate Stability`) %>%
  dplyr::rename(risk_debt = 
                  `Risk Points for Debt Service as a % of XGS`) %>%
  dplyr::rename(risk_CA = 
                  `Risk Points for Current Account as % of XGS`) %>%
  dplyr::rename(risk_liq = 
                  `Risk Points for International Liquidity`) %>%
  dplyr::rename(risk_agg_fin = `Aggregate Financial Risk`)
 )

(fin_risk_tidy_summ <- fin_risk_tidy %>%
    dplyr::group_by(Year) %>%
    dplyr::summarise(risk_agg_min = 
                       min(risk_agg_fin, na.rm = T),
                     risk_agg_max = 
                       max(risk_agg_fin, na.rm = T),
                     risk_agg_med = 
                       median(risk_agg_fin, na.rm = T),
                     risk_agg_mean = 
                       mean(risk_agg_fin, na.rm = T),
                     risk_agg_std = 
                       sd(risk_agg_fin, na.rm = T),
                     risk_agg_iqr =
                       IQR(risk_agg_fin, na.rm = T)
                     )
    
)

ggplot(data = filter(fin_risk_tidy, 
                     Country %in% c("United Kingdom",
                                  "United States",
                                  "China",
                                  "India")
                     ),
       mapping = aes(x = Year, 
                     y = risk_agg_fin,
                     color = Country)) +
  geom_point() +
  geom_line(mapping = aes(linetype = Country))


```

### Gathering

"Gathering" involves moving column names into a "key" column, gathering the column values into a
single value column.

This is used to "gather" data from the wide format, to the long format.

```{r data_gather}

gdppc %>% head(.)

col_yr_1 <- "1960 [YR1960]"
col_yr_end <- "2016 [YR2016]"

(gdppc_tidy <- gdppc %>%
    dplyr::select(-c(`Series Name`, #why backticks ``?
                     `Series Code`,
                     `Country Code`
                     )
                  ) %>%
    dplyr::rename(Country = `Country Name`) %>%
    tidyr::gather(col_yr_1:col_yr_end,
                  key = "Year",
                  value = "GDP_per_capita"
                  ) %>%
    dplyr::arrange(Country)

  )


```



### Spreading

Spreading is the opposite of gathering. It moves the unique values of a key column into the column names, thus "spreading" the column values across new columns.

Below, we spread the variable `fin_risk` which is in the long format, into the wide format.

```{r data_spread}

(fin_risk_foreign <- fin_risk_tidy %>% 
  dplyr::select(c(Country, Year, risk_foreign))
 )

(fin_risk_spread <- fin_risk_foreign %>%
  tidyr::spread(key = Country, 
                value = risk_foreign
                )
  )

```

Each column now is the aggregate foreign financial risk of a country. This is the wide format.

### Joining

`fin_risk_foreign` and `gdppc_tidy` are both tidy datasets containing information about countries' foreign financial risk variables and GDP per capita respectively. What if we wish to merge the two datasets into one, so that we have countries' risk as well as GDP per capita profiles?

This can be achieved by the so-called "mutating join" family of functions from `dplyr`. There are mainly four types of joins: 

1. `left_join(x, y)` merges `y` using matching values for `x`.
2. `right_join(x, y)` merges `x` using matching values for `y`.
3. `inner_join(x, y)` merges only matching rows.
4. `full_join(x, y)` merges fully, retaining all values.

Missing matches are filled with `NA` values.

```{r data_mutating_joins}

gdppc_tidy$Year <- 1960:2016 #reformatting dates

# Left Joining
(data_join_left <- dplyr::left_join(fin_risk_foreign, gdppc_tidy, 
                                   by = c("Country", "Year")
                                   ) %>%
    dplyr::arrange(Country)
  )

# Right Joining
(data_join_right <- dplyr::right_join(fin_risk_foreign, gdppc_tidy, 
                                   by = c("Country", "Year")
                                   ) %>%
    dplyr::arrange(Country)
  )

# Inner Joining
(data_join_inner <- dplyr::inner_join(fin_risk_foreign, gdppc_tidy, 
                                   by = c("Country", "Year")
                                   ) %>%
    dplyr::arrange(Country)
  )

# Full Joining
(data_join_full <- dplyr::full_join(fin_risk_foreign, gdppc_tidy, 
                                   by = c("Year", "Country")
                                   ) %>%
    dplyr::arrange(Country)
  )


```