---
title: "Data Wrangling with `dplyr`"
author: Abhinav Anand

output:
  pdf_document:
    keep_tex: true

fontsize: 11pt
documentclass: article
geometry: margin = 1.75in

linkcolor: blue
urlcolor: red
citecolor: magenta

citation_package: natbib
bibliography: Working_Paper.bib

header-includes:
   - \linespread{1.3}


---

```{r setup, eval=T, message=FALSE, warning=T, include=FALSE, echo=F}

library(tidyverse)
library(rmarkdown)
library(knitr)
library(xml2)
library(gapminder)

knitr::opts_chunk$set(echo = T, 
                      warning = T, 
                      message = F, 
                      eval = T, 
                      include = T
                      )



```

For data wrangling, the package `dplyr` is employed. For downloading, type `install.packages("dplyr")` or equivalently type `install.packages("tidyverse")`


```{r load_gapminder}

(data_gapminder <- gapminder::gapminder)

#what is the format: wide or long?
#what is a tibble?
```

### Notes

1. **Data Types**: R has many in-built data types. Examples:
    i. `fct`: "factor": categorical data which can assume finite levels, say A, B, C etc.
    ii. `dbl`: "double": real numbers, say 3.671, 4.00, 10.122482929 etc.
    iii. `int`: "integer": integers, say 3, 10, -9, 0 etc.
    iv. `chr`: "character": say, "FMC", "Term 4" etc.
    v. `lgl`: "logical": $\{0, 1\}\equiv \{\text{T, F}\}$
    vi. `date`, and many more

2. **Tibbles**: Tibbles are essentially data frames, but slightly altered to work better in  tidyverse. (Compare `head(data_frame_name)` versus `tibble_name`.)

# `dplyr()`: The Main Verbs

1. `filter()`: Extract rows
2. `select()`: Extract columns
3. `arrange()`: Order rows
4. `mutate()`: Create new columns (= variables)
5. `summarise()`: Compute summary statistics

The syntax for all five verbs is similar. The first argument is the data frame, followed by the action to be performed using the variable name.

## Filter

Let us observe the state of the world in 1952 and the contrast and compare with that in 2007.

```{r filter_rows}

(data_1952 <- data_gapminder %>%
  dplyr::filter(year == 1952) #extract the rows for year 1952
 ) #note == as opposed to =

(data_2007 <- data_gapminder %>%
  dplyr::filter(year == 2007) #extract the rows for year 2007
 )

```

## Select

Let us also focus on two variables---GDP/capita and life expectancy. We extract both for years 1952 and 2007.

```{r select_columns}

(data_1952_gdppc <- data_1952 %>%
  dplyr::select(country, year, gdpPercap)
 )
(data_2007_gdppc <- data_2007 %>%
  dplyr::select(country, year, gdpPercap)
 )
(data_1952_life_exp <- data_1952 %>%
  dplyr::select(country, year, lifeExp)
 )
(data_2007_life_exp <- data_2007 %>%
  dplyr::select(country, year, lifeExp)
 )

# = dplyr::select(-c(continent, pop, gdpPercap))
```

`dplyr::rename()` is a wrapper function for `select()` which renames the variable in consideration and keeps all other variables intact.

## Arrange

Usage of `arrange()` orders (from first to last) entries on the basis of a variable. 

### Question: Is the set of richest countries the same in 1952 and 2007?

```{r arrange_gdppc}

(data_1952_rich <- data_1952_gdppc %>%
  dplyr::arrange(desc(gdpPercap)) #note the use of desc()
 )

(data_2007_rich <- data_2007_gdppc %>%
  dplyr::arrange(desc(gdpPercap)) #note the use of desc()
 )

```

### Which countries display the highest life expectancy pre-2000?

```{r arrange_life_pre00}

(data_life_pre00 <- data_gapminder %>%
  dplyr::select(country, year, lifeExp) %>%
  dplyr::filter(year <= 2000) %>%
  dplyr::arrange(desc(lifeExp))
 )

```

### Post-2000?

```{r arrange_life_post00}

(data_life_post00 <- data_gapminder %>%
  dplyr::select(country, year, lifeExp) %>%
  dplyr::filter(year > 2000) %>%
  dplyr::arrange(desc(lifeExp))
 )

```


## Mutate

Let's define a new variable called "total GDP" which is the product of the GDP/capita and the total population. To compute it and include it in the list of variables we can use the verb `dplyr::mutate()`.



```{r mutate_GDP_tot}

(data_GDP_tot <- data_gapminder %>%
   dplyr::mutate(GDP_total = pop*gdpPercap/10^9) #in USD billions
 )

```


In general, for `mutate()` to work well, the function must take a vector of values as input and return a vector with the same number of values as output. A short list of functions that can be used with `mutate()` are: 

1. Arithmetic operators: `+`, `-`, `*`, `/`, `^`
2. Logs: `log()`, `log2()`, `log10()`
3. Cumulative aggregates: `cumsum()`, `cumprod()`, `cummin()`, `cummax()`, `cummean()` etc.

and many more.

### Question: Which countries have the highest total GDP in 1952 and 2007?

```{r GDP_tot_most}

(data_GDP_tot %>% 
   dplyr::filter(year == 1952) %>%
   dplyr::arrange(desc(GDP_total))
 )

(data_GDP_tot %>% 
   dplyr::filter(year == 2007) %>%
   dplyr::arrange(desc(GDP_total))
 )
```

### Question: Which are the five smallest economies in 2007 (by total GDP)?

```{r GDP_tot_top_5}

(data_GDP_tot %>% 
   dplyr::filter(year == 2007) %>%
   dplyr::arrange(GDP_total) %>%
   dplyr::filter(rank(GDP_total) <= 5)
 )
```


## Summarise

The function `summarise()` (or equivalently `summarize()`) can be used to compute summary statistics. Here is an example, where we summarize the variable life expectancy for the continent Europe.

```{r summ_life_exp_eur}

(life_exp_summ_eur <- data_gapminder %>%
  dplyr::filter(continent == "Europe") %>%
  dplyr::summarise(average = mean(lifeExp),
                   med = median(lifeExp),
                   std = sd(lifeExp),
                   variance = var(lifeExp),
                   iqr = IQR(lifeExp)
                   )
 )


```

### Grouped Summaries

### Question: What are continent-wise summary statistics?

```{r summ_cont}

(summ_life_exp <- data_gapminder %>%
  dplyr::group_by(continent) %>%
  dplyr::summarise(average = mean(lifeExp),
                   med = median(lifeExp),
                   std = sd(lifeExp),
                   variance = var(lifeExp),
                   iqr = IQR(lifeExp)
                   )
 )

ggplot(data = summ_life_exp, 
       aes(x = continent, y = average)) +
  geom_bar(stat = "identity")


(summ_year_life_exp <- data_gapminder %>%
  dplyr::group_by(year) %>%
  dplyr::summarise(average = mean(lifeExp),
                   med = median(lifeExp),
                   std = sd(lifeExp),
                   variance = var(lifeExp),
                   iqr = IQR(lifeExp)
                   )
 )

ggplot(data = summ_year_life_exp, 
       aes(x = year, y = average)) + 
  geom_point()



```

One can also perform grouped mutates and grouped filters.

