---
title: "More Tips and Tricks"
author: Abhinav Anand, IIMB
date: '`r format(Sys.time(), "%Y/%m/%d")`' #current date

output:
  pdf_document:
    keep_tex: true

fontsize: 12pt
documentclass: article
geometry: margin = 1.5in

linkcolor: blue
urlcolor: red
citecolor: magenta

citation_package: natbib
bibliography: Working_Paper.bib

header-includes:
   - \linespread{1.3}


---

```{r setup, eval=T, message=FALSE, warning=T, include=FALSE, echo=F}

library(tidyverse); theme_set(theme_bw())
library(rmarkdown)
library(knitr)
library(xml2)
library(gapminder)


knitr::opts_chunk$set(echo = T, 
                      warning = T, 
                      message = F, 
                      eval = T, 
                      include = T,
                      fig.height=3.5, 
                      fig.width=3.5,
                      fig.align = 'center'
                      )
```

# Scripts in R

Scripts are files (ending in `.R`) that contain sequences of instructions that we can command a machine to follow. Consier a typical empirical project: we need to read data stored in some type of file; then clean and tidy it, then process it and use descriptive statistics and plots to delineate its features etc. All this can be achieved by storing a sequence of instructions in a script file. Data can be read using the the `read_csv()` function, processing can be done using the package `dplyr` etc.

# Linear regression in R

Generally a linear model takes the following form: 

$$
y = \beta_0 + \beta_1x_1 + \hdots + \beta_mx_m + u
$$
where $u_{n\times 1}$ is the error term. This setup corresponds to an overdetermined linear system of equations leading to a least squares solution:

$$
\hat{\beta} = (X^{\top} X)^{-1} X^{\top}y 
$$

where the explanatory matrix $X_{m\times n}$ contains independent variables $x_1,\hdots,x_m$ as column vectors of size $n\times 1$.

One of the strengths of R is the flexibility and support it offers for linear regression modeling. In order to illustrate it more fully, let us consider data for India in the `gapminder` dataset.

```{r data_Ind_GDP}

data_Ind <- gapminder::gapminder %>% 
  dplyr::filter(country == "India")

ggplot(data_Ind, aes(year, gdpPercap)) +
  geom_point() +
  geom_line() 

```

We see that there has been a large increase in GDP per capita in India. A similar trend is observed for life expectancy:

```{r data_Ind_lifeexp}

ggplot(data_Ind, aes(year, lifeExp)) +
  geom_point() +
  geom_line() 

```

What about the relationship between the two? For example, (all else equal) does GDP per capita of India explain the life expectancy trends observed?

```{r data_Ind_scatter}

ggplot(data_Ind, aes(gdpPercap, lifeExp)) +
  geom_point() 

```

This suggests that the two variables share a positive relation. We can try to check this by means of a linear regression in the following way:

$$
\text{life exp} = \beta_0 + \beta_1 \text{(gdp percap)} + u
$$

In order to implement this step in R is via the following:

```{r data_Ind_lm}

lm_formula <- lifeExp ~ gdpPercap

lm_life_gdppc <- lm(data = data_Ind, formula = lm_formula)

summary(lm_life_gdppc)

ggplot(data_Ind, aes(gdpPercap, lifeExp)) +
  geom_point() +
  geom_smooth(method = "lm")


```

What is this object `lm_life_gdppc`? What is its structure? We can quickly check by accessing its contents:

```{r data_Ind_lm_object}

names(lm_life_gdppc)

```

What about some subset of data, say the period before 1990?

```{r data_Ind_lm_subset}

lm(filter(data_Ind, year <= 1990), formula = lm_formula) %>% 
  summary()

ggplot(filter(data_Ind, year <= 1990), aes(lifeExp, gdpPercap)) +
  geom_point() +
  geom_smooth(method = "lm")

```



## Nonlinear relationships

The plot between the dependent and independent variable suggest a nonlinear relationship. Can we test this simply? Let's consider the following modification:

$$
\text{life exp} = \beta_0 + \beta_1 \text{(gdp percap)}^2 + u
$$

In general, R can accommodate independent variables involving mathematical operators in a regression equation with the function `I()`.

```{r data_Ind_lm_quad}

lm_formula_quad <- lifeExp ~ I(gdpPercap)^2

lm_life_gdppc_quad <- lm(data = data_Ind, formula = lm_formula_quad)

summary(lm_life_gdppc_quad)

ggplot(data_Ind, aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  stat_smooth(method = "lm", 
              formula = y ~ poly(x, 2), #polynomial order 2
              size = 0.8,
              linetype = "dashed"
              )

# What about higher order polynomials?
ggplot(data_Ind, aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  stat_smooth(method = "lm", 
              formula = y ~ poly(x, 3), #polynomial order 3
              size = 0.8,
              linetype = "dotdash"
              )

```

Are visually better fits also evidence of better underlying models? This is a hard question to answer in general---all else equal we prefer models that are parsimonious (have fewer explanatory variables).

# Functional Programming in R

Another very powerful feature of R is its support for functional programming, which in general, involves applying functions to arrays, dataframes, lists etc.

For example, how should one compute the mean across rows of a matrix?

```{r func_prog}

df <- data.frame(C_1 = rnorm(10, 0, 1), 
                 C_2 = rnorm(10, 1, 2),
                 C_3 = rnorm(10, 2, 3)
                 )

head(df)

# One way to solve the problem
rmean_1 <- rowMeans(df)
print(rmean_1)

# Another more 'functional' way
func_mean <- function(vec)
{
  return(mean(vec, na.rm = T))
}

# Apply function on rows
rmean_2 <- apply(df, 1, func_mean) 
print(rmean_1)

# Apply function on columns
rmean_3 <- apply(df, 2, func_mean)
print(rmean_3)

```

Note how to use the `apply()` function. We `apply()` the function over rows or columns or other dimensions. In general that's the philosophy of the `apply()` family of functions, which includes functions `lapply()` (list-apply) and `sapply()` (simplify-apply) etc. The function `lapply` returns a list and `sapply` a vector (if possible). In both cases the first argument is a list (or dataframe) and the second argument is the name of a function.

What is a list? It's essentially a more general version of a dataframe and can contain not only dissimilar data types but also, say dataframes within them.

```{r func_prog_apply}

temp_list <- list(a = runif(10),
                  b = "Happy birthday",
                  c = data.frame(x = rnorm(10, 0, 1)),
                  d = sample(letters, 7, replace = TRUE)
                  )
str(temp_list) #structure of the list

# 1apply() is used to apply the same function to each
# "cell" of the list
lapply(temp_list, is.numeric) #check if each cell is numeric

# contrast with sapply()
sapply(temp_list, is.numeric)

```

## The `map()` family from `purrr`

The `map` function does the exact same operation as `apply` but is consistent with the output format type. For example `map()` returns a list, `map_dbl()` returns a double type vector, `map_int()` returns an integer type vector etc. As with `read_csv`, the `map` family improves upon the base R code by being faster and more consistent.

```{r func_prog_map}

map(df, mean)

map_dbl(df, median)

# Also compare this
z <- list(x = 1:3, y = 4:5)
map_int(z, length) #names are preserved

```

# Advanced tricks: Nesting and multiple models

We return to the `gapminder` dataset to look at GDP per capita around the world.

```{r nest_mult_mod}

gapminder %>%
  ggplot(., aes(year, gdpPercap, group = country)) +
  geom_line(alpha = 0.3) +
  ylim(0, 50000)

```

Overall, it looks like GDP per capita has been steadily improving though some countries see some declines.

## Nested data frame

This is best understood by means of examples.

```{r nest_mult_mod_intro}

head(gapminder)

# What nesting does
nest_country <- gapminder %>%
  dplyr::group_by(country, continent) %>%
  tidyr::nest()

nest_country

# Accessing country-level data
nest_country$data[[1]] #Afghanistan

```

Now we apply previously discussed ideas from functional programming to such nested dataframes. We want to check if the trend has been rising for each country. This can be achieved by applying a trend-computing function to each dataframe in the nested object.

```{r nest_mult_mod_intro_2}

# Trend computing function for GDP per capita
lm_gdppc_year <- function(data_frame)
{
  temp <- lm(data = data_frame,
             formula = gdpPercap ~ year) #trend
  return((temp))
}

# Update the nested dataframe by incorporating trends
nest_country <- nest_country %>%
  dplyr::mutate(trend = map(data, lm_gdppc_year))
# Equivalently: map(nest_country$data, lm_gdppc_year)

nest_country$trend[[1]] #Afghanistan

# We can also do the standard procedures
nest_country %>%
  dplyr::filter(continent == 'Asia')

```

### Unnesting

The `unnest()` function undoes the nesting:

```{r nest_mult_mod_unnest}

tidyr::unnest(nest_country, data)

```

### Evaluating model fits

The `glance()` function from the package `broom` (included in `tidyverse`) is useful

```{r model_fit}

nest_country %>%
  dplyr::mutate(model_fit = map(trend, broom::glance)) %>%
  tidyr::unnest(model_fit)

```