---
title: "Introduction to ggplot"

output: 
  pdf_document:
    keep_tex: true
  
documentclass: article
geometry: margin = 1.5in

linkcolor: blue
urlcolor: red
citecolor: magenta

citation_package: natbib
bibliography: Working_Paper.bib

header-includes:
   - \linespread{1.3}


---

```{r library_declaration, eval=T, message=FALSE, warning=T, include=FALSE, echo=F}

library(tidyverse)
library(rmarkdown)
library(knitr)
library(xml2)
library(gapminder)

```

# Setup

The following discussion assumes you have donwloaded R and RStudio. Additionally, the package suite
`tidyverse()` which includes the package `ggplot2()` needs to be included.

1. For downloading R, visit <https://cran.r-project.org/>
2. For downloading RStudio visit <https://www.rstudio.com/>
3. For downloading `ggplot2()`, type `install.packages("ggplot2")` or equivalently for `tidyverse()`
type `install.packages("tidyverse")`

# Introduction to ggplot

`ggplot()` stands for ``grammar of graphics'' [@Wilkinson:2005], [@Wickham:2010]. 

This idea will be further explored by the means of data from the package `gapminder()`. To install, type `install.packages("gapminder")`.

```{r R2, eval=T, message=F, warning=T, include=T, echo=T}

temp_data <- gapminder::gapminder 
# <- versus =
# Why gapminder::gapminder?
```



# References