---
title: "More Tips and Tricks"
author: Abhinav Anand, IIMB
date: '`r format(Sys.time(), "%Y/%m/%d")`' #current date

output:
  pdf_document:
    keep_tex: true

fontsize: 12pt
documentclass: article
geometry: margin = 1.5in

linkcolor: blue
urlcolor: red
citecolor: magenta

citation_package: natbib
bibliography: Working_Paper.bib

header-includes:
   - \linespread{1.3}


---

```{r setup, eval=T, message=FALSE, warning=T, include=FALSE, echo=F}

library(tidyverse)
library(rmarkdown)
library(knitr)
library(xml2)
library(gapminder)

knitr::opts_chunk$set(echo = T, 
                      warning = T, 
                      message = F, 
                      eval = T, 
                      include = T,
                      fig.height=3.5, 
                      fig.width=3.5,
                      fig.align = 'center'
                      )
```

# Scripts in R


# Linear estimation in R

Generally a linear model takes the following form: 

$$
y = \beta_0 + \beta_1x_1 + \hdots + \beta_mx_m + u
$$
where $u_{n\times 1}$ is the error term. This setup corresponds to an overdetermined linear system of equations leading to a least squares solution:

$$
\hat{\beta} = X^{\top} X^{-1} X^{\top}y 
$$

where the explanatory matrix $X_{m\times n}$ contains independent variables $x_1,\hdots,x_m$ as column vectors of size $n\times 1$.

One of the strengths of R is the flexibility and support it offers for linear regression modeling. In order to illustrate it more fully, let us consider data for India in the `gapminder` dataset.

```{r data_Ind_GDP}

data_Ind <- gapminder::gapminder %>% 
  dplyr::filter(country == "India")

ggplot(data_Ind, aes(year, gdpPercap)) +
  geom_point() +
  geom_line() +
  theme_bw()

```

We see that there has been a large increase in GDP per capita in India. A similar trend is observed for life expectancy:

```{r data_Ind_lifeexp}

data_Ind <- gapminder::gapminder %>% 
  dplyr::filter(country == "India")

ggplot(data_Ind, aes(year, lifeExp)) +
  geom_point() +
  geom_line() +
  theme_bw()

```

What about the relationship between the two? For example, (all else equal) does GDP per capita of India explain the life expectancy trends observed?

```{r data_Ind_scatter}

ggplot(data_Ind, aes(gdpPercap, lifeExp)) +
  geom_point() +
  theme_bw()

```

This suggests that the two variables share a positive relation. We can try to check this by means of a linear regression in the following way:

$$
\text{life exp} = \beta_0 + \beta_1 \text{(gdp percap)} + u
$$