---
title: "Data Wrangling with `dplyr()`"
author: Abhinav Anand

output:
  pdf_document:
    keep_tex: true

  
documentclass: article
geometry: margin = 1.75in

linkcolor: blue
urlcolor: red
citecolor: magenta

citation_package: natbib
bibliography: Working_Paper.bib

header-includes:
   - \linespread{1.3}


---

```{r setup, eval=T, message=FALSE, warning=T, include=FALSE, echo=F}

library(tidyverse)
library(rmarkdown)
library(knitr)
library(xml2)
library(gapminder)

data_gapminder <- gapminder::gapminder 

```

```{r temp, eval=T, message=F, warning=T, include=T, echo=T}

gapminder
#what is the format? wide or long?
```

### Tibbles

It prints differently because it’s a tibble. Tibbles are data frames, but slightly tweaked to work better in the tidyverse

1. `fct`
2. `dbl`
3. `int`

Other types: `chr`, `lgl`, `date` etc.

# `dplyr()`: The Main Verbs

Pick observations by their values (filter()).
Reorder the rows (arrange()).
Pick variables by their names (select()).
Create new variables with functions of existing variables (mutate()).
Collapse many values down to a single summary (summarise()).
These can all be used in conjunction with group_by() which changes the scope of each function from operating on the entire dataset to operating on it group-by-group. These six functions provide the verbs for a language of data manipulation.

All verbs work similarly:

The first argument is a data frame.

The subsequent arguments describe what to do with the data frame, using the variable names (without quotes).

The result is a new data frame.

## Filter


## Arrange



## Select


## Mutate


## Summarise


## Grouped Mutate